Combining predicate transformer semantics for effects

### 2020-08-28 msfp2020/msfp2020.tex
Update slides
Wouter's suggestions; better theme

\documentclass{beamer}

\usepackage{fontspec}
\usepackage{minted}
\usepackage{xspace}

\setmainfont{TeX Gyre Heros}
\setmonofont{Noto Sans Mono}
\usetheme{Madrid}

\let \parskip=\baselineskip
\setminted{fontsize=\footnotesize}

\newcommand{\Agda}[1]{\texttt{\footnotesize #1}\xspace}
\newcommand{\Free}{\Agda{Free}}
\newcommand{\match}{\Agda{match}}
\newcommand{\Nondet}{\Agda{Nondet}}
\newcommand{\Regex}{\Agda{Regex}}
\newcommand{\Set}{\Agda{Set}}

\title[Combining predicate transformer semantics]{Combining predicate transformer semantics for effects: a case study in parsing regular languages}
\author{Anne Baanen \and Wouter Swierstra}
\date{September 1, 2020}

@@

\titlepage
\begin{frame}[fragile]{Algebraic effects}

@@

@@

A predicate transformer for commands \Agda{C} and responses \Agda{R}
is a function from postconditions of type \Agda{R -> Set} to preconditions of type \Agda{C -> Set}.
If \Agda{R} depends on \Agda{C}, this becomes:

@@

\begin{frame}[fragile]{Parsing regular expressions}
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty         xs       = Op Fail λ()
match Epsilon       Nil      = Pure tt
match Epsilon       (_ :: _) = Op Fail λ()
match (Singleton c) xs       =
    if xs = [c] then Pure c else Op Fail λ()

match (l | r)       xs       = Op Choice (λ b ->
    if b then Inl <$> match l xs else Inr <$> match r xs)

match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}
Looks straightforward so far...

\pause
... until we reach the last case:
\vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
    -- Error: does not terminate
\end{minted}

@@

\begin{frame}[fragile]{Parsing regular expressions}
For now, we will write:
\vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = Op Fail λ()
\end{minted}
\vspace{\baselineskip}

@@

A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt2} satisfies more postconditions than \Agda{pt1}:
\vspace{-\baselineskip}
\begin{minted}{Agda}
_⊑_ : (pt1 pt2 : (a -> Set) -> Set) -> Set
pt1 ⊑ pt2 = ∀ P -> pt1 P -> pt2 P
\end{minted}

@@

\vspace{\baselineskip}

@@

Predicate transformers are a semantic domain where programs and specifications can be related.
\vspace{-\baselineskip}
\begin{minted}{Agda}
[[_,_]] : (pre : Set) (post : a -> Set) -> (a -> Set) -> Set
[[ pre , post ]] P = pre ∧ ∀ x, post x -> P x
\end{minted}

@@

\begin{frame}[fragile]{Adding effects}
The \Agda{Rec I O} effect represents a recursive function of type \Agda{(i : I) -> O i} calling itself.
The commands are the arguments to the function and the responses are the returned values.
\begin{minted}{Agda}
Rec : (I : Set) (O : I -> Set) -> Sig
Rec I O = mkSig I O
\end{minted}
To specify the semantics of \Agda{Rec}, we need an invariant of type \Agda{(i : I) -> O i -> Set},
specifying which values of type \Agda{O i} can be returned from a call with argument \Agda{i : I}.
\begin{minted}{Agda}
ptRec inv i P = ∀ o -> inv i o -> P o
\end{minted}
\end{frame}

@@

\Agda{Parser} has stateful semantics: to return the next character, we need to keep track of the remaining characters.
The state is the extra \Agda{String} arguments in \Agda{ptParser}.
\begin{minted}{Agda}
ptParser : ⊤ -> (Maybe Char -> String -> Set) -> String -> Set
ptParser _ P Nil = P Nothing Nil
ptParser _ P (x :: xs) = P (Just x) xs
\end{minted}

@@

\vspace{\baselineskip}
\Agda{terminates-in f S n} holds iff \Agda{S} terminates after calling \Agda{f} at most \Agda{n} times.
\vspace{-\baselineskip}

@@

In our paper, we illustrate how techniques from the refinement calculus can be used in functional programming.
A distinguishing characteristic of our approach is modularity:
we add new effects and semantics to the system as we need them.

>>>>

\documentclass[ignorenonframetext,]{beamer}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
%%W: moved up to allow to be overriden
\usetheme[]{metropolis}
\usefonttheme{serif} % use mainfont rather than sansfont for slide text
%End
% \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
%   \usepackage[T1]{fontenc}
%   \usepackage[utf8]{inputenc}
% % \else % if luatex or xelatex
%   \ifxetex
%     \usepackage{mathspec}
%   \else
    \usepackage{fontspec}
  % \fi
  % \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[Scale=0.8]{Open Sans}
    \setsansfont[Scale=0.8]{Open Sans}
    \setmonofont[Mapping=tex-ansi,Scale=0.8]{Droid Sans Mono}
%\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\newif\ifbibliography

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}

%% UU Specific colors
\definecolor{main}{RGB}{60,67,149}
\definecolor{text}{RGB}{30,34,75}
\definecolor{background}{RGB}{255,255,255}

\definecolor{alert}{RGB}{0,159,227}
\definecolor{example}{RGB}{60,67,149}

%\definecolor{uuRed}{RGB}{192,10,53}
%\definecolor{uuYellow}{RGB}{255,205,0}
%\definecolor{darkGrey}{RGB}{64,64,64}
%\definecolor{almostBlack}{RGB}{26,26,26}

%\definecolor{uuCreme}{RGB}{249,230,173}
%\definecolor{uuOrange}{RGB}{221,149,98}
%\definecolor{uuBordeaux}{RGB}{146,30,86}
%\definecolor{uuBrown}{RGB}{99,63,43}
%\definecolor{uuGreen}{RGB}{99,165,147}
%\definecolor{uuBlue}{RGB}{102,135,195}
%\definecolor{uuDarkBlue}{RGB}{23,29,66}
%\definecolor{uuPurple}{RGB}{82,41,128}

\setbeamercolor{normal text}{fg=text, bg=white}
\setbeamercolor{alerted text}{fg=alert}
\setbeamercolor{example text}{fg=text}
\setbeamercolor{frametitle}{bg=main,fg=white}
\setbeamercolor{palette primary}{%
   use=normal text
  ,fg=white
  ,bg=main
  }

\titlegraphic{\hfill\includegraphics[height=1.5cm]{img/vu.png}\includegraphics[height=1.5cm]{img/uulogo.png}}

% Prevent slide breaks in the middle of a paragraph:
\widowpenalties 1 10000
\raggedbottom

\AtBeginPart{
  \let\insertpartnumber\relax
  \let\partname\relax
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \let\insertsectionnumber\relax
    \let\sectionname\relax
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \let\insertsubsectionnumber\relax
  \let\subsectionname\relax
  \frame{\subsectionpage}
}

\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

\usepackage{minted}
\usepackage{xspace}

%\setmainfont{TeX Gyre Heros}
\setmonofont[Scale=0.8]{Noto Sans Mono}
%\usetheme{Madrid}

% Make blocks of code appear a bit more cohesive by putting the lines closer together.
\linespread{0.95}

\newcommand{\Agda}[1]{\texttt{#1}\xspace}
\newcommand{\Free}{\Agda{Free}}
\newcommand{\match}{\Agda{match}}
\newcommand{\Nondet}{\Agda{Nondet}}
\newcommand{\Regex}{\Agda{Regex}}
\newcommand{\Set}{\Agda{Set}}

\title[Combining predicate transformer semantics]{Combining predicate transformer semantics for effects}
\subtitle{A case study in parsing regular languages}
\author{Anne Baanen \hspace{5.1em} Wouter Swierstra}
\institute{Vrije Universiteit Amsterdam \hspace{1em} Utrecht University}
\date{September 1, 2020}

\date{}

@@

\frame{\titlepage}

\begin{frame}[fragile]{Algebraic effects}

@@

In this work, we use a free monad to model effectful programs:

@@

A predicate transformer for commands \Agda{C} and responses \Agda{R}
is a function from postconditions of type \Agda{R -> Set} to preconditions of type \Agda{C -> Set}.\\
If \Agda{R} depends on \Agda{C}, this becomes:

@@

\begin{frame}[fragile]{Parsing regular expressions}

We implement \match as a case distinction.
\begin{minted}{Agda}
match : (r : Regex) -> String -> Nondet (Tree r)
match Empty         xs       = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match Epsilon       Nil      = Pure tt
match Epsilon       (_ :: _) = Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (Singleton c) xs       =
    if xs = [c] then Pure c else Op Fail λ()
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l | r)       xs       = Op Choice (λ b ->
    if b then Inl <$> match l xs else Inr <$> match r xs)
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (l · r) xs = do
    (ys, zs) <- allSplits xs
    (,) <$> match l ys <*> match r zs
\end{minted}
\pause \vspace{-\baselineskip}
\begin{minted}{Agda}
match (r *) xs = match (r · (r *)) xs
\end{minted}
\pause
\alert{Error: \Agda{match (r *) xs} does not terminate}

@@

\begin{frame}[fragile]{Parsing regular expressions}
For now, we will write:
\begin{minted}{Agda}
match (r *) xs = Op Fail λ()
\end{minted}

@@

A predicate transformer \Agda{pt1} \emph{is refined by} \Agda{pt2} if \Agda{pt2} satisfies more postconditions than \Agda{pt1}:

\begin{minted}{Agda}
_⊑_ : (pt1 pt2 : (a -> Set) -> Set) -> Set
pt1 ⊑ pt2 = ∀ P -> pt1 P -> pt2 P
\end{minted}

@@

@@

Predicate transformers are a semantic domain where programs and specifications can be related.
\begin{minted}{Agda}
[[_,_]] : (pre : Set) (post : a -> Set) -> (a -> Set) -> Set
[[ pre , post ]] P = pre ∧ ∀ x, post x -> P x
\end{minted}

@@

\begin{frame}[fragile]{Adding effects}
Inspired by McBride's \textit{Turing-Completeness Totally Free}, we use the \Agda{Rec I O} effect to represent a recursive function of type \Agda{(i : I) -> O i} calling itself.
The commands are the arguments to the function and the responses are the returned values.
\begin{minted}{Agda}
Rec : (I : Set) (O : I -> Set) -> Sig
Rec I O = mkSig I O
\end{minted}
To specify the semantics of \Agda{Rec}, we need an invariant of type \Agda{(i : I) -> O i -> Set},
specifying which values of type \Agda{O i} can be returned from a call with argument \Agda{i : I}.
\begin{minted}{Agda}
ptRec inv i P = ∀ o -> inv i o -> P o
\end{minted}
\end{frame}

@@

\Agda{Parser} has stateful semantics: to return the next character, we need to keep track of the remaining characters.
The state is the extra \Agda{String} arguments in \Agda{ptParser}.
\begin{minted}{Agda}
ptParser : (Maybe Char -> String -> Set) -> String -> Set
ptParser P Nil = P Nothing Nil
ptParser P (x :: xs) = P (Just x) xs
\end{minted}

@@

\Agda{terminates-in f S n} holds iff \Agda{S} terminates after calling \Agda{f} at most \Agda{n} times.

@@

In our paper, we illustrate how techniques from the refinement calculus can be used in functional programming.
They provide a natural and uniform way to reason about effects in the setting of the \Agda{Free} monad.

A distinguishing characteristic of our approach is modularity:
we add new effects and semantics to the system as we need them.
